%\section[Overview]{ Overview
\chapter[Overview]{ Overview
\label{sec:targets-overv}
\footnote{
  $CVS~revision~ $Id: overview.tex,v 1.11 2008/04/18 20:05:18 jpchen Exp $ $ }
\footnote{Authors: J. P. Chen \email{jpchen@jlab.org}}
}

Three types of mutually exclusive target systems have been used in Hall A:

\begin{list}{\arabic{enumi}.~}{\usecounter{enumi}\setlength{\itemsep}{-0.15cm}}
  \item a system of cryo-targets and solid targets;
  \item a waterfall target;
  \item a target of polarized gaseous $^3$He.
\end{list}

%The physics program in Hall A utilizes a number of different
%target systems of varying complexity. 
The set of
cryogenic targets currently operates with liquid hydrogen, liquid
deuterium and gasous helium 3 or helium 4
as target materials.
A variety of solid targets are also provided; BeO, Carbon and
Aluminum are typical but other self supporting materials are available if need arises.
The combination of cryogenic targets and a few solid targets is
the standard configuration. \\

A waterfall target was used during the commissioning of
the hall spectrometers and for hypernuclear experiments. This system also 
requires a special installation. \\

In addition,
there is a large program based on polarized $^3$He. This
is a special installation and hence is not available at the same
time as the cryogenic target system.\\

Each of these systems is discussed in following chapters.
% ===========  CVS info
% $Header: /group/halla/analysis/cvs/tex/osp/src/targets/overview.tex,v 1.11 2008/04/18 20:05:18 jpchen Exp $
% $Id: overview.tex,v 1.11 2008/04/18 20:05:18 jpchen Exp $
% $Author: jpchen $
% $Date: 2008/04/18 20:05:18 $
% $Name:  $
% $Locker:  $
% $Log: overview.tex,v $
% Revision 1.11  2008/04/18 20:05:18  jpchen
% new target overview
%
% Revision 1.10  2004/12/15 00:14:15  gen
% minor update of the labels
%
% Revision 1.9  2003/12/17 03:59:49  gen
% authorized personnel tables unfied
%
% Revision 1.8  2003/12/13 06:23:39  gen
% Septum added. Name tables. Polishing
%
% Revision 1.7  2003/12/05 07:35:03  gen
% WT2 modified. Polishing
%
% Revision 1.6  2003/11/21 18:26:21  gen
% polishing
%
% Revision 1.5  2003/11/17 07:19:13  gen
% CVS record added
%
% Revision 1.4  2003/09/23 19:56:06  jpchen
% overview of targets
%
% Revision 1.3  2003/06/06 17:09:04  gen
% Revision printout changed
%
% Revision 1.2  2003/06/05 23:30:01  gen
% Revision ID is printed in TeX
%
% Revision 1.1.1.1  2003/06/05 17:28:27  gen
% Imported from /home/gen/tex/OSP
%
%  Revision parameters to appear on the output
